Screening Sinkhorn Algorithm for Regularized Optimal Transport

### 2019-05-21 tex/v3/main.tex
update sec 1&3

\section{Conclusion}
The paper introduces a novel efficient approximation of the Sinkhorn divergence
based on a screening strategy. Screening some of the Sinkhorn dual variables
has been made possible by defining a novel constrained dual problem and by 
carefully analyzing its optimality conditions. From the latter, we derived some
sufficient conditions depending on the ground cost matrix, that some dual variables are smaller than a given threshold. Hence, we need just to solve a restricted
dual Sinkhorn problem using an off-the-shelf LBFGS algorithm. We also provided
some theoretical guarantees of the quality of the approximation with respect to
the number of variables that has been screened. Numerical experiments show 
the behaviour of our \emph{Screenkhorn} algorithm and computational time gain it can
achieve when integrated in some complex machine learning pipelines.

>>>>

\section{Conclusion}
The paper introduces a novel efficient approximation of the Sinkhorn divergence
based on a screening strategy. Screening some of the Sinkhorn dual variables
has been made possible by defining a novel constrained dual problem and by 
carefully analyzing its optimality conditions. From the latter, we derived some
sufficient conditions depending on the ground cost matrix, that some dual variables are smaller than a given threshold. Hence, we need just to solve a restricted
dual Sinkhorn problem using an off-the-shelf L-BFGS-B algorithm. We also provide
some theoretical guarantees of the quality of the approximation with respect to
the number of variables that has been screened. Numerical experiments show 
the behaviour of our \emph{Screenkhorn} algorithm and computational time gain it can
achieve when integrated in some complex machine learning pipelines.

### 2019-05-21 tex/v3/section_3_algorithm.tex
update

Pseudocode of our proposed algorithm is shown in Algorithm~\ref{screenkhorn}. %~\cite{nocedal1980,nocedal2006numerical}
It is worth to note that Screenkhorn algorithm uses only the restricted parts $K_{(I_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})},$ $K_{(I_{\varepsilon,\kappa}, J^\complement_{\varepsilon,\kappa})},$ and $K_{(I^\complement_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})}$ of the Gibbs kernel $K$ for calculating the objective function $\Psi_{\varepsilon, \kappa}$ and its gradient, in contrast to Sinkhorn algorithm which performs alternating updates of all rows and columns of $K.$
Furthermore, Screenkhorn consists of two steps: the first one is an initialization providing the active sets $I_{\varepsilon,\kappa}$, $J_{\varepsilon,\kappa}$. 
The second is a constrained L-BFGS-B for the stacked variable $\theta=(u_{I_{\varepsilon,\kappa}},v_{J_{\varepsilon,\kappa}}).$ 
L-BFGS-B handles box constraints on variables, and it becomes more efficient when these box bounds are carefully determined for problem~\eqref{screen-sinkhorn_second_def}. 
The following proposition (proof in supplementary material) expresses these bounds that are pre-calculated in the initialization step of Screenkhorn.
\begin{proposition}
\label{prop:bounds_of_usc_and_vsc}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal pair solution of problem~\eqref{screen-sinkhorn_second_def} and $K_{\min} = \min\limits_{i\in I_{\varepsilon,\kappa},j \in J_{\varepsilon,\kappa}}K_{ij}$. Then,
one has
\begin{equation}
\label{bound_on_u}
\frac \varepsilon\kappa \vee \frac{\min_{i \in I_{\varepsilon,\kappa}}\mu_i}{\varepsilon (m- m_b) + \varepsilon \vee \frac{\max_{j\in J_{\varepsilon,\kappa}} \nu_j}{n\varepsilon\kappa K_{\min}} m_b} \leq e^{u^{\text{sc}}_i} \leq \frac \varepsilon\kappa\vee \frac{\max_{i \in I_{\varepsilon,\kappa}} \mu_i}{m\varepsilon K_{\min}},
\end{equation}
and
\begin{equation}
\label{bound_on_v}
\varepsilon\kappa \vee \frac{\min_{j \in J_{\varepsilon,\kappa}}\nu_j}{\varepsilon(n- n_b) + \varepsilon \vee \frac{\kappa\max_{i\in I_{\varepsilon,\kappa}} \mu_i}{m\varepsilon K_{\min} } n_b} \leq e^{v^{\text{sc}}_j} \leq \varepsilon\kappa \vee \frac{\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\varepsilon K_{\min} }
\end{equation}
for all $i\in I_{\varepsilon,\kappa}$ and $j\in J_{\varepsilon,\kappa}$.
\end{proposition}
\LinesNotNumbered
\begin{algorithm}[htbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{Screenkhorn$(C,\eta,\mu,\nu,n_b,m_b)$}
\label{screenkhorn}

@@

\nl  $u^{(0)}\gets \log(\varepsilon\kappa^{-1}) \mathbf 1_{n_b} ;$\\
\nl  $v^{(0)} \gets \log(\varepsilon\kappa) \mathbf 1_{m_b} ;$\\
\nl  $\theta^{(0)} \gets \text{stack}[u^{(0)}, v^{(0)}];$\\
\nl   $\theta \gets \text{L-BFGS-B}(\theta^{(0)}, \underline{\theta}, \bar{\theta});$\\
\nl   $\theta_u \gets (\theta_1, \ldots, \theta_{n_b})^\top, \theta_v \gets(\theta_{n_b+1}, \ldots, \theta_{n_b+m_b})^\top;$\\
\nl   {$u^{sc}_i \gets (\theta_u)_i$ if $i \in I_{\varepsilon,\kappa}$ and $u_i \gets \log(\varepsilon\kappa^{-1})$ if $i \in I^\complement_{\varepsilon,\kappa};$}\\
\nl   {$v^{sc}_j \gets (\theta_v)_j$ if $j \in J_{\varepsilon,\kappa}$ and $v_j \gets \log(\varepsilon\kappa)$ if $j \in J^\complement_{\varepsilon,\kappa};$}\\
\nl   \Return{$B(u^{\text{sc}},v^{\text{sc}})$.}
\end{algorithm}

>>>>

Pseudocode of our proposed algorithm is shown in Algorithm~\ref{screenkhorn}. %~\cite{nocedal1980,nocedal2006numerical}
It is worth to note that Screenkhorn algorithm uses only the restricted parts $K_{(I_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})},$ $K_{(I_{\varepsilon,\kappa}, J^\complement_{\varepsilon,\kappa})},$ and $K_{(I^\complement_{\varepsilon,\kappa}, J_{\varepsilon,\kappa})}$ of the Gibbs kernel $K$ for calculating the objective function $\Psi_{\varepsilon, \kappa}$ and its gradient, in contrast to Sinkhorn algorithm which performs alternating updates of all rows and columns of $K.$
Furthermore, Screenkhorn consists of two steps: the first one is an initialization providing the active sets $I_{\varepsilon,\kappa}$, $J_{\varepsilon,\kappa}$. 
The second is a constrained L-BFGS-B for the stacked variable $\theta=(u_{I_{\varepsilon,\kappa}},v_{J_{\varepsilon,\kappa}}).$ 
L-BFGS-B handles box constraints on variables, and it becomes more efficient when these box bounds are carefully determined for problem~\eqref{screen-sinkhorn_second_def}. 
The following proposition (proof in supplementary material) expresses these bounds that are pre-calculated in the initialization step of \emph{Screenkhorn}.
\begin{proposition}
\label{prop:bounds_of_usc_and_vsc}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal pair solution of problem~\eqref{screen-sinkhorn_second_def} and $K_{\min} = \min\limits_{i\in I_{\varepsilon,\kappa},j \in J_{\varepsilon,\kappa}}K_{ij}$. Then,
one has
\begin{equation}
\label{bound_on_u}
\frac \varepsilon\kappa \vee \frac{\min_{i \in I_{\varepsilon,\kappa}}\mu_i}{\varepsilon (m- m_b) + \varepsilon \vee \frac{\max_{j\in J_{\varepsilon,\kappa}} \nu_j}{n\varepsilon\kappa K_{\min}} m_b} \leq e^{u^{\text{sc}}_i} \leq \frac \varepsilon\kappa\vee \frac{\max_{i \in I_{\varepsilon,\kappa}} \mu_i}{m\varepsilon K_{\min}},
\end{equation}
and
\begin{equation}
\label{bound_on_v}
\varepsilon\kappa \vee \frac{\min_{j \in J_{\varepsilon,\kappa}}\nu_j}{\varepsilon(n- n_b) + \varepsilon \vee \frac{\kappa\max_{i\in I_{\varepsilon,\kappa}} \mu_i}{m\varepsilon K_{\min} } n_b} \leq e^{v^{\text{sc}}_j} \leq \varepsilon\kappa \vee \frac{\max_{j \in J_{\varepsilon,\kappa}} \nu_j}{n\varepsilon K_{\min} }
\end{equation}
for all $i\in I_{\varepsilon,\kappa}$ and $j\in J_{\varepsilon,\kappa}$.
\end{proposition}
\LinesNotNumbered
\begin{algorithm}[htbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{Screenkhorn$(C,\eta,\mu,\nu,n_b,m_b)$}
\label{screenkhorn}

@@

\nl  $u^{(0)}\gets \log(\varepsilon\kappa^{-1}) \mathbf 1_{n_b} ;$\\
\nl  $v^{(0)} \gets \log(\varepsilon\kappa) \mathbf 1_{m_b} ;$\\
\nl  $\theta^{(0)} \gets \texttt{stack}[u^{(0)}, v^{(0)}];$\\
\nl   $\theta \gets \text{L-BFGS-B}(\theta^{(0)}, \underline{\theta}, \bar{\theta});$\\
\nl   $\theta_u \gets (\theta_1, \ldots, \theta_{n_b})^\top, \theta_v \gets(\theta_{n_b+1}, \ldots, \theta_{n_b+m_b})^\top;$\\
\nl   {$u^{sc}_i \gets (\theta_u)_i$ if $i \in I_{\varepsilon,\kappa}$ and $u_i \gets \log(\varepsilon\kappa^{-1})$ if $i \in I^\complement_{\varepsilon,\kappa};$}\\
\nl   {$v^{sc}_j \gets (\theta_v)_j$ if $j \in J_{\varepsilon,\kappa}$ and $v_j \gets \log(\varepsilon\kappa)$ if $j \in J^\complement_{\varepsilon,\kappa};$}\\
\nl   \Return{$B(u^{\text{sc}},v^{\text{sc}})$.}
\end{algorithm}

### 2019-05-21 tex/v3/section_5_numeric_expes.tex
update sec 5

We have implemented our Screenkhorn in Python and used the LBFGS of
Scipy. Regarding the machine-learning based comparison, we have based our code
on the ones of POT~\citep{flamary2017pot} and just replaced the Sinkhorn function call with a Screenkhorn one. We have considered the POT's default Sinkhorn stopping criterion parameters and for Screenkhorn, the LBFGS algorithm is stopped when the 
largest component of the projected gradient is smaller than $10^{-6}$, when the number of iterations of number of objective function evaluations reach $10^{5}$. For all applications, we have set $\eta=1$ unless otherwise specified.

>>>>

We have implemented our Screenkhorn in Python and used the L-BFGS-B of
Scipy. Regarding the machine-learning based comparison, we have based our code
on the ones of POT~\citep{flamary2017pot} and just replaced the Sinkhorn function call with a Screenkhorn one. We have considered the POT's default Sinkhorn stopping criterion parameters and for Screenkhorn, the L-BFGS-B algorithm is stopped when the 
largest component of the projected gradient is smaller than $10^{-6}$, when the number of iterations of number of objective function evaluations reach $10^{5}$. For all applications, we have set $\eta=1$ unless otherwise specified.
